Making multiple-locus association mapping on a genome-wide scale routine

### 2017-10-27 AM+draft.tex
27Oct17 Methods - about to work on Lmm-Lasso section

\subsubsection{Reducing the dimension of the model}
In practice, estimating the parameters of (\ref{eq2}) can be demanding, computationally. 
The vector $a_{-S}$ has $L-s$ random effects where in modern genome-wide association studies, 
$L$, the number of snp, can be extremely large.  An alternative model is given by 
Verbyla. 
They show how to reformulate (\ref{eq2}) to be a model with a random effect with only $n$ elements
\begin{equation}
\label{eq3}
u_g = \sum_{k=1}^s  m_{S_k} a_{S_k} + (M_{-S} M_{-S}^T)^{1/2} a^*_{-S}
\end{equation}
where $a^* \sim N(0, \sigma_a^2 I^{(n \; \times \;  n)})$, and 
$(M_{-S} M_{-S}^T)^{1/2}$ can be calculated via single value decomposition (ref).  
Although it may not be obvious, the two models are equivalent, 
having identical variance structures. Yet, the computational cost of model (\ref{eq3}) compared to 
model (\ref{eq2}) is much less due to the random term having only $n$ effects needing estimating.

Verbyla go on to show how to recover $\widetilde{a}$ from estimates from model (3) with 
\begin{equation}
\widetilde{a} = \left [ M_{-S}^T (M_{-S} M_{-S}^T)^{-1/2} \right ] \widetilde{a}^*
\end{equation}
where its variance matrix is
\begin{equation}
\label{eq4}
\textrm{var}(\widetilde{a}) = M_{-S}^T (M_{-S} M_{-S}^T)^{-1/2} \textrm{var}(\widetilde{a}^*) (M_{-S} M_{-S}^T)^{-1/2} M_{-S}
\end{equation}
These values are needed in order to calculate the score statistic $t_j^2$ for identifying the snp in strongest association with the trait. 
Fortunately, when calculating $t_j^2$, only the diagonal elements of the variance matrix are needed which simplifies the  calculation. 
of (\ref{eq4}).

>>>>

\subsubsection{Reducing the dimension of the model}
In practice, estimating the parameters of (\ref{eq2}) can be demanding, computationally. 
The vector $a_{-S}$ has $L-s$ random effects where in modern genome-wide association studies, 
$L$, the number of snp, can be extremely large.  An alternative model is given by 
Verbyla \cite{verbyla2012rwgaim,verbyla2014whole}. 
They show how to reformulate (\ref{eq2}) to be a model with a random effect with only $n$ elements
\begin{equation}
\label{eq3}
u_g = \sum_{k=1}^s  m_{S_k} a_{S_k} + (M_{-S} M_{-S}^T)^{1/2} a^*_{-S}
\end{equation}
where $a^* \sim N(0, \sigma_a^2 I^{(n \; \times \;  n)})$, and 
$(M_{-S} M_{-S}^T)^{1/2}$ can be calculated via single value decomposition (ref).  
Although it may not be obvious, the two models are equivalent, 
having identical variance structures. Yet, the computational cost of model (\ref{eq3}) compared to 
model (\ref{eq2}) is much less due to the random term having only $n$ effects needing estimating.

Verbyla \cite{verbyla2012rwgaim,verbyla2014whole} go on to show how to recover $\widetilde{a}$ from estimates from model (3) with 
\begin{equation}
\widetilde{a} = \left [ M_{-S}^T (M_{-S} M_{-S}^T)^{-1/2} \right ] \widetilde{a}^*
\end{equation}
where its variance matrix is
\begin{equation}
\label{eq4}
\textrm{var}(\widetilde{a}) = M_{-S}^T (M_{-S} M_{-S}^T)^{-1/2} \textrm{var}(\widetilde{a}^*) (M_{-S} M_{-S}^T)^{-1/2} M_{-S}
\end{equation}
These values are needed in order to calculate the score statistic $t_j^2$ for identifying the snp in strongest association with the trait. 
Fortunately, when calculating $t_j^2$, only the diagonal elements of the variance matrix are needed which simplifies the  calculation. 
of (\ref{eq4}).

\subsection{Comparison Packages and Methods}

The performance of the Eagle package is compared against seven other genome-wide association packages.
Five of these packages, BigRR, glmnet, FaST-LASSO, MMLM, r2VIM implement methods that have been applied to the multiple locus analysis of data from genome-wide association studies. 
The other two packages, GEMMA, and FaST-LMM implement single-locus association mapping methods. 
The packages 
differ in their features (summarized in Table X).  The packages also differ in their underlying methods (summarized below).

\subsubsection{Multiple locus methods}

\textbf{BigRR:} The BigRR package implements generalised ridge regression  for
the analysis of genetic data \cite{shen2013novel}.  It is a regularisation method in that it produces regression solutions 
that have higher bias than ordinary least squares but reduced variance for improved prediction accuracy. The underlying 
model can be formulated as  $y = X \beta + M u + e$ where $u^{(L \times 1)}$ is a vector of genetic effects, and each genetic 
effect is normally distributed with its own variance such that $u_l \sim N(0, \sigma^2_l); l=1,2, \ldots, L$.  Here, the 
amount of shrinkage is allowed to vary across the genetic (snp) effects which differs to 
standard ridge regression where the shrinkage is constant. 
Generalised ridge regression is apt at estimating the effect sizes of snp on a genome-wide scale but  obtaining their 
statistical significance is not straightforward. This makes using the BigRR package for association mapping a challenge. 
Shen \cite{shen2013novel} suggest permutation as a way of identifying the genomic regions of interest 
but we instead employ stability selection (described below).

\textbf{glmnet:}  The glmnet package is a general purpose statistical package that implements elastic net, a regularisation 
method for generalised linear models  \cite{zou2005regularization}.  The model used for association mapping is 
$y = X \beta + \sum_{l=1}^L m_l \theta_l + e$ where $\theta_l; l=(1,2, \ldots, L)$ is the regression coefficient for the $l$th snp. 
It's penalty function for regularisation is a mixture of $\ell_1$ (lasso) and $\ell_2$ (ridge regression).  This allows elastic net to avoid some of the
limitations of lasso (ref) while still being able to perform shrinkage and variable selection, simultaneously.  The glmnet package 
implements a modification of the LARS algorithm \cite{efron2004least} for finding the optimal regularisation (solution) path. However, when 
applied to the analysis of data from genome-wide association studies we found this yielded findings containing a high proportion of 
false positives.  We instead fixed the shrinkage parameter to a suitable value 
and employed stability selection to estimate the significance of the snp regression coefficients, empirically (see below).

  MMLM

  r2VIM

  LMM-lasso

Single locus methods

gemma

Fast-LMM
